---
title: "figure 2"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
date: "2026-03-03"
---

```{r, message=F, warning=F}
rm(list=ls())
library(rprojroot)
library(dplyr)
root <- rprojroot::find_rstudio_root_file()
```



## define active enhancer 

* overlap ATAC seq peaks and H3k27ac peaks, and not within 2kb of TSS


```{r, echo=F, message=F, warning=F}
library(readxl)

 ################ these peak windows are from tianxiang 
cre_h3k27ac_peaks=read_excel("C:\\han\\Projects\\ShareShengtong\\ShareShengtong\\EpiBmal1Cerebellum\\Draft\\Files\\Figure2-PanelA_cre_h3k27ac_chip-seq_regions.xlsx", col_names=F)
colnames(cre_h3k27ac_peaks)[1:3]=c("chr", "start", "end")

flx_h3k27ac_peaks=read_excel("C:\\han\\Projects\\ShareShengtong\\ShareShengtong\\EpiBmal1Cerebellum\\Draft\\Files\\Figure2-PanelA_flx_h3k27ac_chip-seq_regions.xlsx", col_names=F)
colnames(flx_h3k27ac_peaks)[1:3]=c("chr", "start", "end")

########## ATAC peaks called from new sample 
atac_peaks=read_excel(file.path(root, "../Data/Figures_Generation_Data/figure_1A.xlsx")) # this is union peaks from both cre and flx 

cre_atac_peaks=as_tibble(read.table(file.path(root, "../Data/Novogene/PeakWindows/Cere_C_merged_peaks.narrowPeak")))

flx_atac_peaks=as_tibble(read.table(file.path(root, "../Data/Novogene/PeakWindows/Cere_K_merged_peaks.narrowPeak")))
```



```{r, message=F, warning=F}
########### load libraries and set up 
library(tidyverse)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm10.knownGene) # For Mouse mm10 TSS


# 1. Convert ATAC peaks (format chr1:start:end)   # this is union peak 
atac_gr <- atac_peaks %>%
  separate(peak_window, into = c("chr", "start", "end"), sep = ":", remove = FALSE) %>%
  mutate(start = as.numeric(start), end = as.numeric(end)) %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

### this is cre peaks 
colnames(cre_atac_peaks)[1:3] <- c("chr", "start", "end")
# Convert to GRanges
cre_atac_gr <- cre_atac_peaks %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

### this is flx peaks 
colnames(flx_atac_peaks)[1:3] <- c("chr", "start", "end")
# Convert to GRanges
flx_atac_gr <- flx_atac_peaks %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)




# 2. Convert H3K27ac peaks
cre_h3k27ac_gr <- cre_h3k27ac_peaks %>%
  mutate(
    start = as.numeric(start),
    end   = as.numeric(end)
  ) %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

flx_h3k27ac_gr <- flx_h3k27ac_peaks %>%
  mutate(
    start = as.numeric(start),
    end   = as.numeric(end)
  ) %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)


# Get TSS for mm10
genes_gr <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
tss_gr <- resize(genes_gr, width = 1, fix = "start")
# . Expand it to 2kb (1000bp on each side)
# This is much safer than punion/shift
tss_2kb_zone <- tss_gr + 1000

# Filter ATAC windows that are NOT in the TSS zone (Distal regions)
cre_atac_distal_gr <- atac_gr[!overlapsAny(cre_atac_gr, tss_2kb_zone)]
flx_atac_distal_gr <- atac_gr[!overlapsAny(flx_atac_gr, tss_2kb_zone)]


cre_active_enhancers_gr <- subsetByOverlaps(cre_atac_distal_gr, cre_h3k27ac_gr)
cat("Found", length(cre_active_enhancers_gr), "cre active distal enhancers.\n")

flx_active_enhancers_gr <- subsetByOverlaps(flx_atac_distal_gr, flx_h3k27ac_gr)
cat("Found", length(flx_active_enhancers_gr), "flx active distal enhancers.\n")

```

```{r, message=F, warning=F}
library(TxDb.Mmusculus.UCSC.mm10.knownGene) # For Mouse mm10 TSS
library(org.Mm.eg.db)
########## find enhancer genes 
get_enhancer_genes <- function(enhancer_gr, max_distance = 100000) {
  # 1. Get mm10 Gene/TSS annotations
  genes_gr <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
  tss_gr <- resize(genes_gr, width = 1, fix = "start")
  
  # 2. Find the nearest TSS for each enhancer
  # distanceToNearest returns a hits object with distance metadata
  near_idx <- distanceToNearest(enhancer_gr, tss_gr)
  
  # 3. Extract and combine data
  # queryHits: index of enhancer | subjectHits: index of gene
  res <- as.data.frame(enhancer_gr[queryHits(near_idx)]) %>%
    mutate(
      gene_id = tss_gr$gene_id[subjectHits(near_idx)],
      distance_to_tss = mcols(near_idx)$distance
    )
  
  # 4. Map Entrez IDs to Gene Symbols
  res$gene_symbol <- mapIds(org.Mm.eg.db, 
                            keys = res$gene_id, 
                            column = "SYMBOL", 
                            keytype = "ENTREZID", 
                            multiVals = "first")
  
  # 5. Clean up and filter by distance (default 100kb)
  final_df <- res %>%
    as_tibble() %>%
    filter(distance_to_tss <= max_distance) %>%
    dplyr::select(gene_symbol, distance_to_tss, gene_id, seqnames, start, end, everything()) %>%
    arrange(distance_to_tss)
  
  message(paste("Mapped", nrow(final_df), "enhancers to genes within", max_distance/1000, "kb."))
  
  return(final_df)
}
cre_enhancer_gene=get_enhancer_genes(cre_active_enhancers_gr)
flx_enhancer_gene=get_enhancer_genes(flx_active_enhancers_gr)
```



```{r, message=F, warning=F}
# draw plot 2(A)
cre_numbers=c(nrow(cre_h3k27ac_peaks), length(cre_active_enhancers_gr), length(unique(cre_enhancer_gene$gene_symbol)))
flx_numbers=c(nrow(flx_h3k27ac_peaks), length(flx_active_enhancers_gr), length(unique(flx_enhancer_gene$gene_symbol)))

hist_data=data.frame(count=c(flx_numbers, cre_numbers), variable=rep(c("Peaks", "Active Enhancers", "Enhancer Genes"), 2), sample=rep(c("Flx", "Cre"), each=3))
hist_data$sample <- factor(hist_data$sample, levels = c("Flx", "Cre"))  # this ensures Cre on the top and Flx at the bottom 



# Set the order of the variables to match the image (Peaks at the top)
hist_data$variable <- factor(hist_data$variable, 
                             levels = c("Enhancer Genes", "Active Enhancers", "Peaks"))

#  Draw the plot
ggplot(hist_data, aes(x = sample, y = count, fill = variable)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  # Add text labels
  geom_text(aes(label = scales::comma(count)), 
            position = position_dodge(width = 0.8), 
            hjust = -0.1,    # Push text slightly to the right of the bar end
            size = 5,        # Font size of the numbers
            fontface = "bold") +
  coord_flip() + # Makes it horizontal
  scale_fill_manual(values = c("Peaks" = "#7030A0",            # Purple
                               "Active Enhancers" = "#0070C0", # Dark Blue
                               "Enhancer Genes" = "#00B0F0")) + # Light Blue
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + # Starts at exactly 0
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal() +
  theme(
    # --- Axis Text Customization ---
    axis.text.y = element_text(face = "bold", size = 18, color = "black"),
    axis.text.x = element_text(size = 14, color = "black", face = "bold"),
    
    # --- Legend Font Size Customization ---
    legend.position = "bottom",
    legend.text = element_text(size = 16, face = "bold"), # Increases legend text size
    legend.key.size = unit(1.2, "cm"),                   # Makes the color boxes larger
    
    # --- Clean up Grid ---
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "gray80"), 
    plot.margin = margin(10, 20, 10, 10) # Adds a bit of margin around the whole plot
  )

```



1. **Active Enhancer Definition: ATAC peak windows intersect with  H3K27ac peak windows, except 2kb proximity of TSS**.

2. ATAC peaks: use Cre and Flx specific narrowPeak windows called from 2 combined replicates  

3. **enhancer genes: the nearest gene for each enhancer**






```{r, message=F, warning=F}
library(GenomicRanges)
library(dplyr)
library(ggvenn)

# 1. Set your cutoff
length_cutoff <- 1000  # Try 10, then 1000, then 5000

# 2. Find overlaps and filter by width
overlaps <- findOverlaps(cre_active_enhancers_gr, flx_active_enhancers_gr)
overlap_widths <- width(pintersect(cre_active_enhancers_gr[queryHits(overlaps)], 
                                   flx_active_enhancers_gr[subjectHits(overlaps)]))

# 3. Identify pairs that meet the cutoff
valid_hits <- overlaps[overlap_widths >= length_cutoff]
length(valid_hits)

# 4. CREATE ENTIRELY UNIQUE IDs (Crucial Step)
#  generate random unique strings so NO peak matches by default
cre_unique_ids <- paste0("Cre_Peak_", seq_along(cre_active_enhancers_gr))
flx_unique_ids <- paste0("Flx_Peak_", seq_along(flx_active_enhancers_gr))

# 5. FORCE MATCHES ONLY FOR VALID HITS
# If they pass the filter, give them the SAME ID in both lists
match_ids <- paste0("Match_", seq_along(valid_hits))
cre_unique_ids[queryHits(valid_hits)] <- match_ids
flx_unique_ids[subjectHits(valid_hits)] <- match_ids

# 6. Build the list for ggvenn
venn_list <- list(
  "Cre-Gained" = cre_unique_ids,
  "Cre-Lost"   = flx_unique_ids
)

# 7. Plot
ggvenn(venn_list, 
       fill_color = c("#CCEEFF", "#66CCEE"),
       stroke_size = 1.5, 
       set_name_size = 0, 
       text_size = 6,
       show_percentage = TRUE) +
  annotate("text", x = -1.2, y = 1.2, label = "Cre-Gained", size = 6, fontface = "bold") +
  annotate("text", x = 1.2, y = 1.2, label = "Cre-Lost", size = 6, fontface = "bold") +
  annotate("text", x = 0, y = -1.1, label = "Unchanged", size = 6, fontface = "bold") +
  labs(title = paste0("Active Enhancers (Cutoff: ", length_cutoff, "bp)")) +
  theme_void()+
  theme(
    plot.title = element_text(hjust = 0.5,     # Centers the title
                             size = 22,        # Large size
                             face = "bold",    # Bolds the title
                             vjust = 2),       # Adds vertical spacing
    plot.margin = margin(20, 20, 20, 20)
  )
```





* Cre-Gained: active enhancers only in Cre; Cre-Lost: active enhancers only in Flx; Unchanged: active enhancers in both cre and flx with overlap greater than cutoff.   











```{r, message=F, warning=F}
library(ggvenn)
library(ggplot2)

# 1. Define sets - Ensure Strengthened is first (Left) and Weakened is second (Right)
venn_list <- list(
  "Uniq-Cre-Gained" = cre_enhancer_gene$gene_symbol,
  "Uniq-Cre-Lost"     = flx_enhancer_gene$gene_symbol
)

# 2. Plot with specific label placements
ggvenn(venn_list, 
       fill_color = c("#CCEEFF", "#66CCEE"),
       stroke_size = 1.5, 
       set_name_size = 0,      # Hide default set names so we can place them manually
       text_size = 6,
       show_percentage = T) +
  # Label for "Strengthened" (Top Left)
  annotate("text", x = -1.2, y = 1.2, label = "Uniq-Cre-Gained", 
           size = 6, fontface = "bold") +
  # Label for "Weakened" (Top Right)
  annotate("text", x = 1.2, y = 1.2, label = "Uniq-Cre-Lost", 
           size = 6, fontface = "bold") +
  # Label for "Unchanged" (Bottom Middle of intersection)
  annotate("text", x = 0, y = -1.1, label = "Unchanged", 
           size = 6, fontface = "bold") +
  labs(title = "Enhancer Genes") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold", vjust = 2),
    plot.margin = margin(10, 10, 10, 10)
  )
```
